---
title: "lab_02"
author: "derek willis"
date: "2024-08-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

For this lab we'll be working with a slice of Maryland voter registration data: records of people who have registration dates this year through Aug. 7. Let's load it.

```{r}
md_voters_2024 <- read_rds("data/lab_02.rds")
```


```{r}
head(md_voters_2024)
summary(md_voters_2024)
colnames(md_voters_2024)
glimpse(md_voters_2024)
```

## Examine Data

Take a look at the data, getting familiar with what it contains. Most of the columns should be self-explanatory, but there are two date columns: STATE_REGISTRATION_DATE and COUNTY_REGISTRATION_DATE. The former represents the date the person registered as a Maryland voter; the latter is when they registered in their current county. They may be the same, but if a voter moves from one county to another they will differ. This is important because if the two dates are the same, that's a pretty good indication that this is a *new* voter, someone who has no history of registering in Maryland. You'll need to keep that definition in mind.

## Answer questions

**Q1** How many voters appear to be *new* registrants since July 21, when President Joe Biden announced he would leave the race, based on the information above? Create a dataframe to hold the only those voters. Which counties have seen the most? Take a look at county population figures here and compare the two in your answer: https://worldpopulationreview.com/us-counties/maryland.

**A1 There is a total of 17076 new registrants since July 21. Montgomery County had the most with 4847 new registrants. When clicking on the link, it shows that Montgomery County has the highest population in Maryland for 2024. It also showed a rise in population since 2020.**



```{r}
md_voters_2024 |>
  filter(STATE_REGISTRATION_DATE == COUNTY_REGISTRATION_DATE, STATE_REGISTRATION_DATE >
  as.Date("2024-07-21")) |>
  mutate(new_voters = "yes") |>
  count(COUNTY, sort = "TRUE") |>
  arrange(desc(n))
```
  

**Q2** A better question is whether the new voters you found in Q1 are an outlier compared to other weeks this year. Using the original dataframe with 2024 registrants, use `mutate()` and lubridate to produce week numbers from the COUNTY_REGISTRATION_DATE (Lubridate has a function that turns a date into a week number. Use that to answer this question). Save that to the original dataframe. Then calculate how many voters are in each week. Describe the distribution of voters groups in a sentence - are the weeks from July 21-August 7 different from the rest, and if so, how?

**A2 After comparing the data, I noticed that during these three weeks the numbers were much higher making them an outlier. ChatGPT helped me fix some of my errors since I was struggling finding the outlier. **

```{r}
# Load necessary packages
library(dplyr)
library(lubridate)

md_voters_2024 <- md_voters_2024 |> 
  mutate(week_number = week(COUNTY_REGISTRATION_DATE))
  weekly_counts <- md_voters_2024 |> 
  group_by(week_number) |> 
  summarise(total_registrations = n()) |> 
  arrange(week_number)
  summary(weekly_counts, total_registrations)
  print(weekly_counts)
```

**Q3** Using lubridate, show which month saw the largest number of registrations. Does the answer help shed more light on Q2?
**A3 The month that had the largest number of registrations was July. Yes, this helped me better understand what month had the highest registration total thanks to the month number being next to the total number whereas the one above just went by week number.** 

```{r}
md_voters_2024 |>
  mutate(month = month(COUNTY_REGISTRATION_DATE)) |>  
  group_by(month) |> 
  summarise(total_registrations = n()) |>  
  arrange(desc(total_registrations)) 
```

**Q4** Maryland has, officially, five regions: https://www.visitmaryland.org/article/maryland-regions. Modify the original dataframe again to add a `region` column based on the `COUNTY` value using case_when. Are you going to have to do this for every Maryland county plus Baltimore City? Yes, you are. Which regions and parties account for the most new voters?

**A4 The Capital Region, sepcifically Montgomery County and the democratic party account for the most new voters. I used ChatGPT to help modify my code and fix my errors especially when I wasn't understanding why I had NA written in a few of the columns but it explained what I did wrong and told me how I could fix it.** 

```{r}
 md_voters_2024 |> 
  mutate(region = case_when(
    COUNTY %in% c("Garrett", "Allegany", "Washington") ~ "Western Maryland",
    COUNTY %in% c("Montgomery", "Frederick", "Prince George's") ~ "Capital Region",
    COUNTY %in% c("Baltimore County", "Baltimore City", "Carroll", "Harford", "Howard", "Anne Arundel") ~ "Central Maryland",
    COUNTY %in% c("Charles", "Calvert", "Saint Mary's") ~ "Southern Maryland",
    COUNTY %in% c("Cecil", "Kent", "Queen Anne's", "Talbot", "Caroline", "Dorchester", "Wicomico", "Worcester", "Somerset") ~ "Eastern Shore",
  )) |>
  
group_by(region, COUNTY, PARTY) |>
summarise(new_voters = n()) |>
arrange(desc(new_voters))
```

**Q5** What's an additional question you could ask of this data?

**A5 How does different political affiliations correlate with different counties acorss the five regions of Maryland?** 
