---
title: "Class Reference"
author: "Derek Willis"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example, this sets the working directory to a directory called data_journalism_2024_fall inside a directory called jour472 located inside your home directory (the ~ is equivalent of /Users/[youruser] on a Mac).

### How to install and load the tidyverse

```{r}
install.packages("tidyverse")
library(tidyverse)
```

```{r}
# This is a comment inside of a codeblock. Comments doesn't run when we run the codeblock.
# Adding this command is a good way to start your class reference notebook.
library(tidyverse)
```

```{r}
# obs is short for observation, which is another way of saying rows or records.
```

```{r}


#print will take a string and return it in the terminal 
print("Hello world")

```

```{r}
 #|>   the pipe operator; interpret this as "and then do this" for mac: cmd-shift-m 
```

```{r}
#lubridate deals with month, day and year and allows functions for each of them and allows for the expansion of mathematical operations that can be performed when using date and time objects. Lubridate makes it easier to work with date and times that R already does but it also allows to complete tasks that R can't do. 
```

```{r}
#head( ) shows the first few rows of a dataset, typically the first 6 rows by default
#summary( ) provides the summary of each column in a dataset 
#colnames( ) returns or sets the names of the columns in a dataset 
#glimpse( ) allows for a quick overview of a dataset showing column names, data types, and a preview of values 
```

```{r}
#*100 turns the decimal into a percentage which makes it easier to read
#arrange(desc) allows for the highest number of a specific column to be displayed at the top. 
```

```{r}
# `str_to_upper` that will convert a character column into all uppercase.
```

```{r}
#Filtering creates a subset of the data based on criteria. Filtering works with rows -- when we filter, we get fewer rows back than we start with.
#Selecting simply returns only the fields named. Selecting works with columns. You will have the same number of records when you are done, but fewer columns of data to work with.
```

```{r}
#The comparison operators in R, like most programming languages, are == for equal to, != for not equal to, \> for greater than, \>= for greater than or equal to and so on.

#Be careful: `=` is not `==` and `=` is not "equal to". `=` is an assignment operator in most languages -- how things get named.
```

```{r}
# OR is additive; AND is restrictive.

#A general tip about using filter: it's easier to work your way towards the filter syntax you need rather than try and write it once and trust the result. Each time you modify your filter, check the results to see if they make sense. This adds a little time to your process but you'll thank yourself for doing it because it helps avoid mistakes.
```

