---
title: "Reverse Engineering Project"
author: "Brinkley Smith"
date: "March 12 2025"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are reverse engineering the story, [CEOs gave heavily during legislative session, exposing loophole in NC's fundraising ban](https://www.wral.com/in-session-campaign-donations/17716104/)
## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
# turn off sci notation
options(scipen=999)
library(tidyverse)
library(lubridate)
library(refinr)
library(dplyr)
library(janitor)
```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

```{r}
# Load required data
donors <- read_csv("data/donors0103-0701.csv")
# Path to data should be loaded from folder "data" i.e. read_csv("data/name_of_data.csv")
```

```{r}
donors <-donors |>
  mutate(
    amount = str_remove_all(amount, "\\$"),
    amount = str_remove_all(amount, ",")
  )
glimpse(donors)
```


```{r}
# Clean required data and prepare for analysis if needed. 
donors <- donors |>
  mutate(date = mdy(date))|>
  mutate(amount =as.numeric(amount))|>
  mutate(zip = str_sub(zip, start = 1L, end = 5L)) |>
  mutate(committee_zip = str_sub(committee_zip, start = 1L, end = 5L))
```

```{r}
glimpse(donors)
```


## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

* **Sentence text**: “Add in the week before session began, and the two businessmen gave more than $87,000 to Republican legislators and the North Carolina Republican Party, a key cog in the campaign machine this year as the two major parties fight for control of the General Assembly in November's elections.”

* **Analysis summary**: The results of my analysis allowed for the grouping of the contributions of George Lewis Strobel II and Robin E Delmer. The amounts from each of them were presented too and it exceeded the $87,000. So yes, I was able to confirm the findings. 



```{r}
donors <- donors |>
  rename(
    name_clean = "name-clean"
  )

```

```{r}
  donors |>
  group_by(committee_name)|>
  summarise(amount = sum(amount)) |>
  arrange(desc(amount))

```


```{r}
donors |>
  filter(name_clean == "GEORGE LEWIS STROBEL II" | name_clean == "ROBIN E DELMER") |>
  group_by(name_clean) |>
  summarise(amount = sum(amount)) |>
  arrange(desc(amount))
```
#add code here

```{r}
donors |>
  filter(date >= "2018-01-03" & date <= "2018-01-09") |>
  group_by(name) |>
  summarise(amount) |>
  arrange(desc(amount))
```

### Sentence 2

* **Sentence text**: “Senate President Pro Tem Phil Berger's campaign alone raised nearly $222,000 during session. Twenty-three people gave him the maximum allowed:$5,200.”

* **Analysis summary**: We were able to confirm the findings when we filtered for Berger's campaign the dates of the short sessions and the amount given 5200 we got 23 rows. This number proves the sentence above saying that there was 23 individual donations or people that donated the max amount allowed equaling $222,000. 

```{r}
# Put code to reverse engineer sentence here
# legislative short session started May 16/0
berger <- donors |>
  filter(date >= '2018-05-16' & date <= '2018-06-29') |>
  filter(committee_name == "PHILIP E BERGER COMM") |>
  filter(amount == 5200) |>
  group_by(name_clean, date, amount) |>
  summarise(sum(amount))

print(berger)
```


### Sentence 3

* **Sentence text**: “Beer distributors also gave during session. Mark Craig, the president of distributor R.H. Barringer, put more than $30,000 into state legislative campaigns during the six-week period, including the maximum $5,200 to Berger. He was the top individual donor to legislative campaigns during session."

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 4

* **Sentence text**: “Add in donations to the two main political parties and to a handful of other large fundraising vehicles for state legislative races, and the in-session giving topped $3.4 million. Gov. Roy Cooper's main campaign account brought in another $407,000 during session.”

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 5

* **Sentence text**: “A WRAL News analysis of campaign finance filings also turned up $16,000 in donations to legislative campaigns during session from 15 PACs with a lobbyist. Each would appear to be a violation of state fundraising rules, but spokespeople for most of the PACs, or the companies they represent, said there was a simple explanation: The checks were written before session.”

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

-30-