---
title: "lab_03"
author: "Derek Willis"
date: "2024-08-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)

# Load janitor
library(janitor)
```

## Load data

```{r}
# Load MD slice of FEC House candidate contribution data
md_contribs <- read_rds("data/lab_03.rds") |>
  clean_names() |>
  mutate(contributor_city = case_when(
    contributor_city %in% c("BALRIMORE", "BLATIMORE") ~ "BALTIMORE",
    contributor_city %in% c("ANAPOLIS") ~ "ANNAPOLIS",
    contributor_city %in% c("SILVER SPRINGS") ~ "SILVER SPRING",
    contributor_city %in% c("N BETHESDA") ~ "NORTH BETHESDA",
    contributor_city %in% c("LUTHERVILLE TIMO", "TIMONIUM", "LUTHERVILLE-TIMONIUM", "LUTHVLE TIMON", "LUTHERVILLE", "LUTHERVILLE TIMONI") ~ "LUTHERVILLE TIMONIUM",
    TRUE ~ contributor_city
  ))  
```

## Fix header issues

You'll need to fix one issue with the column names using janitor. Write a code block to do that, making sure that you save the results to the same dataframe.

```{r}
# cleaning function
cleaned_md_contribs <- md_contribs |>
  clean_names()

# display the cleaned dataset
cleaned_md_contribs

```


## Answer questions

Q1: Which seven cities in Maryland had the largest *number* of contributions from donors? How many did each have? To answer this question, you'll have to clean the messy "contributor_city" field.

Note: To answer this question, you'll need to see every variation of the "contributor_city" for the top seven cities, which you should identify through some initial analysis. Then find variations of them and change those variations into the most common version.

A1: The top seven cities were Bethesda(1177), Chevy Chase(704), Baltimore(692), Annapolis(492), Potomac(490), Silver Spring(408), and Rockville(358).

```{r}

top_cities <- cleaned_md_contribs |>
  count(contributor_city, sort = TRUE) |>
  slice_head(n = 7)

print(top_cities)

```


Q2: Find the 5-digit ZIP code in Maryland that had the highest total amount of contributions (by dollar value). You will need to use mutate to ensure that you have a zip code column with the first five characters from the existing zip code column.

Where is that zip code and what kind of area is it, demographically? Which committee is the largest recipient of money from that zip code? How many of the top five recipients are Maryland candidates?

A2: The top zip code in Maryland with he highest total amount of contributions was Bethesda in Montgomery County. The demographics are primarily white, middle aged or senior residents. The committee with the largest recipient of money from that zip code is Helmer for Congress. Two out of the top five recipients are Maryland candidates. I used ChatGPT to help modify my code after receiving error code when i used the mutate function.

```{r}
# Ensure ZIP codes are formatted correctly (extract first 5 digits)
cleaned_md_contribs <- cleaned_md_contribs |>
  mutate(contributor_zip = substr(contributor_zip, 1, 5))  

# Find the ZIP code with the highest total contributions
top_zip <- cleaned_md_contribs |>
  group_by(contributor_zip) |>
  summarise(total_contributions = sum(contribution_receipt_amount, na.rm = TRUE)) |> 
  arrange(desc(total_contributions)) |>
  slice_head(n = 1)

print(top_zip)

# Identify the largest recipient from this ZIP code
top_recipient <- cleaned_md_contribs |>
  filter(contributor_zip == top_zip$contributor_zip) |>
  group_by(committee_name) |>
  summarise(total_received = sum(contribution_receipt_amount, na.rm = TRUE)) |>
  arrange(desc(total_received)) |>
  slice_head(n = 1)

print(top_recipient)

# Check how many of the top 5 recipients are Maryland candidates
top_five_recipients <- cleaned_md_contribs |>
  filter(contributor_zip == top_zip$contributor_zip) |>
  group_by(committee_name)|>
  summarise(total_received = sum(contribution_receipt_amount, na.rm = TRUE)) |>
  arrange(desc(total_received)) |>
  slice_head(n = 5)

# Assuming Maryland candidates contain "MD" in their committee names
md_candidates_count <- sum(str_detect(top_five_recipients$committee_name, "MD"))

print(md_candidates_count)


```

Q3: Let's look at top individual donors. Write code to generate a list of top donors using name and city, adding up the total amount of contributions. What does the original data say about the top two donors - who are they and what do they do? Do you see anything in the list of top 10 donors that would make you uncertain about the accuracy of these results?

A3: The top two donors are Mark P. Jacobsen of Bethesda, Maryland and Melanie Jacobsen of Bethesda and Bryantown, Maryland. Mark is a CEO, banking services banker and Melanie is a farmer. Possible duplicates of names. Also having NA could raise accuracy questions for Melanie.

```{r}
# Aggregate contributions by donor name and city
top_donors <- cleaned_md_contribs %>%
  group_by(contributor_name, contributor_city) |>
  summarise(total_contributions = sum(contribution_receipt_amount, na.rm = TRUE)) |>
  arrange(desc(total_contributions))

# Display the top 10 donors
top_donors %>% slice_head(n = 10)


top_two_donors <- top_donors %>% slice_head(n = 2)

cleaned_md_contribs |>
  filter(contributor_name %in% top_two_donors$contributor_name) %>%
  select(contributor_name, contributor_city, contributor_employer, contributor_occupation) %>%
  distinct()


top_two_donors <- c("JACOBSEN, MELANIE", "JACOBSEN, MARK P.")

cleaned_md_contribs %>%
  filter(contributor_name %in% top_two_donors) %>%
  select(contributor_name, contributor_city, contributor_employer, contributor_occupation) %>%
  distinct()


```
