---
title: "data_analysis"
author: "Brinkley Smith"
output: html_document
date: "2025-04-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
options(scipen=999)
library(tidyverse)
library(lubridate)
library(refinr)
library(dplyr)
library(janitor)
```


```{r}
independent_expenditure <- read_csv("data/independent_expenditure_2024.csv")
```

```{r}
clean_independent_expenditure <- independent_expenditure |>
  clean_names() |>
  rename(
    candidate_id = cand_id,
    candidate_name = cand_name,
    spender_id = spe_id,
    spender_name = spe_nam,
    election_type = ele_type,
    candidate_state = can_office_state,
    candidate_district = can_office_dis,
    candidate_office = can_office,
    candidate_party = cand_pty_aff,
    expenditure_amount = exp_amo,
    expenditure_date = exp_date,
    aggregate_amount = agg_amo,
    support_or_oppose = sup_opp,
    purpose = pur,
    payee = pay,
    file_number = file_num,
    amendment_indicator = amndt_ind,
    transaction_id = tran_id,
    image_number = image_num,
    receipt_date = receipt_dat,
    election_year = fec_election_yr,
    previous_file_number = prev_file_num,
    dissemination_date = dissem_dt
  )
```

### wrote to csv to clean in open refine

```{r}
#importing after cleaning in openrefine
clean_independent_expenditure <- read_csv("data/updated_cleaned.csv")

```


There are 23 columns and 72,971 rows after we cleaned the data we had 1,761 rows. This data represents all of the independent expenditure spending by candidates/committees from 2023-2024. The sheer amount of data might be limiting because there could be errors in the data. It also accounts for all types of races, so to compare certain types of expenditures would require a lot of filtering. Our questions might be hard to answer in terms of finding average spending because of the amount of data. Additionally, comparing geographical spending compared to other years will be hard because the dataset lacks geographical information.
We're gonna need to turn the dates numerical to analyze them better. There also seems to be spelling errors after we ran clean names. We'll need to find definitions for the candidate ID, spe_id, can_office_dis, sup_opp, pur, amndt_ind, and image_num. The FEC has a github which gives many descriptions, so we'll need to input those definitions to make the data clearer.

##Question 1: Which presidential candidates have spent the most during their campaign? Who has spent the least? Answer: Jon Tester is the candidate that has spent the most during the campaign. There are eight candiates though that are tied for the least amount of money spent on their campaign. 

```{r}
clean_independent_expenditure |>
  group_by(candidate_name) |>
  summarise(total = sum(expenditure_amount, na.rm = TRUE)) |>
  arrange(desc(total))
```
#Question 2: What are the most common payment purposes for payment in federal campaign finance data? What is the average amount paid per transaction for these top purposes? 
Answer: Phonebanking -> average amount $980.15
Production Costs -> average amount $177,963.6
Media Production -> average amount $9263.04
Advertisement Management Services -> average amount $928.37
Canvassing -> average amount $65,250.19

```{r}
#This is using the cleaned data set to group by donation "purpose" and then summarizing by both the average donation amount using the mean function and also by the total amount per purpose using the sum function.  
top_purposes <- clean_independent_expenditure |>
  filter(!is.na(purpose), purpose != "") |>
  group_by(purpose) |>
  summarise(
    count = n(),
    average_amount = mean(expenditure_amount, na.rm = TRUE),
    total_amount = sum(expenditure_amount, na.rm = TRUE)
  ) |>
  arrange(desc(count)) |>
  slice_head(n = 10)
#The slice_head(n=10) function filters the arrange() function to limit it to 10 options.

top_purposes
#These results show that PAC's prioritized advertising and voter outreach (things like media production and phone banking) in order to have the greatest possible impact on the election.
```

#Question 3: Which candidates receive the most independent support versus opposition, and how does spending differ between supporting and attacking candidates?
Answer: Jon Tester received the most independent support versus opposition. Democrats tend to spend more money on elections and their campaigns compared to Republicans. 

```{r}
#The main priority of this dataset is seeing which candidates had the most money spent in support or in opposition of them.
summary <- clean_independent_expenditure |>
  filter(!is.na(candidate_name), support_or_oppose %in% c("S", "O")) |>
  group_by(candidate_name, support_or_oppose) |>
  summarise(total_spent = sum(expenditure_amount, na.rm = TRUE), .groups = "drop")

#We used the cleaned_expenditures dataset and created a new summary dataset with it that grouped by candidate_name and support_or_oppose (which differentiates using an "S" or an "O" lettering. We then summarized by total spent on each using the "sum()" function.

#This creates a sub data set from summary that shows the top five total expenditures that were in support of a candidate.
top_supported <- summary |>
  filter(support_or_oppose == "S")|>
  arrange(desc(total_spent)) |>
  head(5)

#This creates a sub data set from summary that shows the top five total expenditures that were in opposition of a candidate.
top_opposed <- summary |>
  filter(support_or_oppose == "O") |>
  arrange(desc(total_spent)) |>
  head(5)


purpose_comparison <- clean_independent_expenditure |>
  filter(!is.na(purpose), support_or_oppose %in% c("S", "O")) |>
  group_by(candidate_name, support_or_oppose) |>
  summarise(total_spent = sum(expenditure_amount, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(total_spent))

#By sorting the data by support_or_oppose and then arranging by total_spent, we can see that the most money spent was in opposition to Jon Tester in Montana's senate race. This makes sense because it was one of the most expensive senate races in history, and Republicans spent a lot of money to make sure that he lost his seat.
purpose_comparison

```


#Question 4: What PAC's donated in both the 2020 and 2024?
Answer: PAC's that donated in both elections are:
- American Crossroads				
- Americans for Constitutional Liberty dba The Conservative Caucus				
- End Citizens United				
- Gun Rights America				
- Indivisible Action				
- MONTANA HUNTERS AND ANGLERS LEADERSHIP FUND				
- MORE JOBS, LESS GOVERNMENT				
- March On PAC				
- Montana Rural Voters (WORC)				
- NEA Advocacy Fund

```{r}
cleaned_2020_expenditures <- read_csv("data/cleaned_2020_expenditures.csv")
```
```{r}
#This is using 2020 data that we plan to join in the next code block. We cleaned this data in OpenRefine and then used the same clean_names() function and column names to rename the columns to something more identifiable and universal.
independent_expenditure_2020 <- cleaned_2020_expenditures |>
  clean_names() |>
  rename(
    candidate_id = cand_id,
    candidate_name = cand_name,
    spender_id = spe_id,
    spender_name = spe_nam,
    election_type = ele_type,
    candidate_state = can_office_state,
    candidate_district = can_office_dis,
    candidate_office = can_office,
    candidate_party = cand_pty_aff,
    expenditure_amount = exp_amo,
    expenditure_date = exp_date,
    aggregate_amount = agg_amo,
    support_or_oppose = sup_opp,
    purpose = pur,
    payee = pay,
    file_number = file_num,
    amendment_indicator = amndt_ind,
    transaction_id = tran_id,
    image_number = image_num,
    receipt_date = receipt_dat,
    election_year = fec_election_yr,
    previous_file_number = prev_file_num,
    dissemination_date = dissem_dt
  )
```

```{r}
#We wanted to know which PAC's donated in both the 2020 and 2024 elections to determine which ones had continued engagement across multiple elections. This is using additional data from the FEC website to grab 2020 election expenditure data. We cleaned it in Open Refine before this.
spenders_2020 <- independent_expenditure_2020|>
  select(spender_name) |>
  distinct()

#This creates a new dataframe called spenders_2020 and then selects the one column spender_name().
spenders_2024 <- clean_independent_expenditure |>
  select(spender_name) |>
  distinct()

#This creates a similar dataframe and selects spender_name, but for 2024. We plan to join them together below.
common_spenders <- inner_join(spenders_2020, spenders_2024, by = "spender_name")

#This uses the inner_join function to join the two created dataframes together by the selected column name "spender_name()."

common_spenders
#This new "common_spenders()" dataframe shows every instance of "spender_name()" that appeared in both the 2020 and the 2024 data, telling us which PAC's donated in both elections.

```



#Question 5: How are Republicans and Democrats spending their money differently? What are each party’s top spending points? 
Answer: Democrats spent a lot of money on television ads, radio and media production. Republicans spent a lot of money on data harvesting. The Democrats top spending point was PLACED MEDIA: TV. The Republicans top spending point was Data Harvesting.

```{r}
clean_independent_expenditure |>
  group_by(candidate_party, expenditure_amount, purpose) |>
  summarise(total = sum(expenditure_amount, na.rm = TRUE)) |>
  arrange(desc(total))
```




