---
title: "lab_01"
author: "Ryan Little"
date: "2025-02-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your reference notebook!

## Load libraries and establish settings

**Task** Import the Tidyverse. Write the code needed to load the library. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
library(tidyverse)
```

## Load Data

In this lab, we will be recreating the first data story I ever did as a data journalism student at Merrill while working in the Howard Center for Investigative Journalism. First, we need to import the court charges data from Okaloosa County, Florida.

**Task** Using the example from the pre_lab, write code to load the CSV file in the data file and assign it to a good variable name. This data is only charges for homeless people.

```{r}
homeless_charges <- read_csv("data/okaloosa-court-charges.csv")

head(homeless_charges)

```

## Examine the data

**Task** Run the codeblock below to use glimpse to get a sense of the column names, types and sample data.

**Q1** How many rows and columns does this dataframe have?

**A1** There are 15,386 rows and 56 columns in this dataframe.

```{r}
glimpse(homeless_charges)

```

## Answer questions

**Q2** Which charges are homeless people most often charged with?

**A2** The charges homeless people are most often charged with is narcotic equip possess or use(991).

```{r}
homeless_charges|>
  group_by(court_statute) |>
  summarise(
    count_contribs = n()
  ) |>
  arrange(desc(count_contribs))
```

**Q3** What charges against homeless people have been charged the least?

**A3** There are several charges that have been charged the least. The first one that appears on the list is adult permints under 16 yoa to possess weapon. 

```{r}
homeless_charges|>
  group_by(court_statute) |>
  summarise(
    count_contribs = n()
  ) |>
  arrange(count_contribs)
```

**Q4** Why do you think there are so many charges identified in **A3** that are tied for the least common? Hint: You can't answer this question with code.

**A4** These charges are very specific charges filed against individuals and are very uncommon, hence why there is only one count for a variety of charges. 

**Q5** Which prosecutor has prosecuted the most charges against homeless people? Is the top prosecutor related to any famous Floridians?

**A5** The prosecutor that has prosecuted the most charges against homeless people is Ginger Bowden Madden. Yes, her father is a former Florida State football coach.

```{r}
homeless_charges|>
  group_by(prosecutor) |>
  summarise(
    count_contribs = n()
  ) |>
  arrange(desc(count_contribs))
```


**Q6** Which judge has overseen the most cases involving homeless people? Hint: Cases and charges are not the same thing.

**A6** The judge that has overseen the most cases involving homeless people is Patricia S Grinsted.

```{r}
homeless_charges|>
  group_by(judge) |>
  summarise(
    count_judge = n_distinct(case_number)
  ) |>
  arrange(desc(count_judge))
```

**Q7** What prosecutor has prosecuted the most cases against a single homeless person?

**A7** Cassie Reed has prosecuted the most cases against a single homeless person.

```{r}
homeless_charges|>
  group_by(prosecutor, defendant) |>
  summarise(
     count_contribs = n()
  ) |>
  arrange(desc(count_contribs))
```

**Q8** We need to create a dataframe that will help us when we are reporting on the ground. Building off our work so far, we need a dataframe that includes the name of every homeless person who was charged with a crime and some important counts about their interactions with the criminal justice system. Create a dataframe that includes the name of each homeless person, the number of charges they have faced, the number of cases they have been involved in, how many dollars they have been assessed in fines, the average fine imposed on them and the median fine imposed upon them. Then sort the data frame so the person with the most charges is at the top. 

**A8**

```{r}
homeless_charges|>
  group_by(defendant) |>
  summarise(
     number_cases= n_distinct(case_number),
    number_charges = n(),
   total_fine = sum(total_assessed),
    mean_fine = mean(total_assessed),
     median_fine= median(total_assessed),
     
  ) |>
  arrange(desc(number_charges))
```

**Q9** Please describe a short reporting plan for how you would use the data you have generated in **A8** to report on the criminal justice system in Okaloosa County, Florida. Who are the people we should talk to? What more should we do with our data analysis to be certain we know what the story is?

**A9* I would like to analyze the trends in data and see if homeless people are being targeted by law enforcement. I'd also look into the fines that these defendants are being charged with. I would like to talk to some of the defendant's that have the most number of charges against them such as Richard Scott Chancey. I want to hear their stories and see how these high fees are impacting them. The prosecutors would be another source to interview and see if I can determine why some individuals are being charged more frequently than others. To see if homeless people are being targeted more I would need to compare data with non-homeless defendants. From there, I could see if those who are homeless are targeted more and fined more. It would be important to also look at repeat offenders and the types of charges that are being filed against an individual.
