---
title: "lab_04"
author: "Derek Willis"
date: "2024-09-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

You'll need to load three packages for this: the tidyverse, lubridate and janitor.

**Task** load these three packages.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(janitor)
library(lubridate)

```

For this lab, we want to investigate spending by the two leading Senate candidates in Maryland during this election, Angela Alsobrooks and Larry Hogan. Our goal is to try and standardize the purpose of the expenditures so that we can compare the two campaigns. We have the data, but it's messy - purpose descriptions are spelled differently and capitalized differently - and we need to clean it up. We'll use a combination of RStudio and OpenRefine to do that.

The basic process here is to start in RStudio, export data so that we can use OpenRefine to clean it up, and then bring it back into RStudio to finish our analysis.

## Load Data

You'll need to load one data set, a CSV file of campaign expenditures located in the data folder called "md_senate_expenses.csv"

**Task** Create a codeblock below, then read the data in and assign it to an appropriate variable name. You'll want to make sure that any date columns are actually date datatypes.

```{r}
# Load the Maryland expenditures data table

# Load data
clean_md_senate_expenses <- read_csv("data/clean_md_senate_expenses.csv")

# Display it
clean_md_senate_expenses
```

## Answer questions

**Q1.** You've been assigned to compare the spending priorities of the two campaigns, but they describe their spending differently using the `purpose` column. Using OpenRefine, create a project using the original CSV file, make a copy of the `purpose` column called `purpose_clean` and then standardize it, focusing on the purposes that appear most often or have the largest total amounts. You can transform the data as you want - making all records the same case, for example, to help. The goal is to be able to group together as many similar purpose values, so you can choose to call all transportation-related expenses the same thing.

How do you know when you're done? It's a judgment call, but you should be close to 100 unique values for `purpose_clean`, and maybe even below that number.

Then, when you are done standardizing `purpose_clean`, export it as a new CSV file to your data folder, *giving it a different name* and read it back into RStudio as a new dataframe.

Then, using that `purpose_clean` column, write code to find the largest amounts of spending on categories by both campaigns.

**A1: The largest amounts of spending from both categories fell under advertising. Alsobrooks spending approximately $4.7 million and Hogan spending approximately $2.5 million on advertising.**

```{r}

# Check the first few rows
head(clean_md_senate_expenses)

# Summarize total spending per category
category_spending <- clean_md_senate_expenses %>%
  group_by(purpose_clean, candidate) %>%
  summarise(total_spent = sum(amount, na.rm = TRUE)) %>%
  arrange(desc(total_spent))

# Display the top 10 spending categories
print(head(category_spending, 10))


```

**Q2.** Let's do some research on the top purposes from A1. Feel free to search the Web to understand those terms better if need be, and then write a description comparing and contrasting the two campaigns' largest categories of spending.

```{r}
clean_md_senate_expenses|>
filter(purpose_clean == 'Advertising')|>
group_by(candidate, organization_name)|>
summarise(total_spent = sum(amount))|>
arrange(desc(total_spent))

```

**A2: Both campaigns focus on spending their money in advertisements. Alsobrooks spent approximately $1.6 million more on advertising  compared to Hogan. Alsobrooks spent the advertising money with a company called Grassroots Media. Hogan spent the most advertising money with a company called M2 Placement.**

**Q3.** Pick one of the top categories for each campaign and write code that shows how that spending has changed over time, if at all. You'll need to use lubridate for this. Write a couple of sentences describing that pattern.

**A3: The spending pattern follows a pre-election ramp-up strategy that shows little spending early on. However, there was a rise in spending going into election season with some gradual declines post peak election.The April 2024 peak suggests that campaigns strategically allocated major resources close to the election for maximum impact.**

```{r}

# Load required libraries
library(dplyr)
library(lubridate)

# Read the clean dataset
df <-  read_csv("data/clean_md_senate_expenses.csv")

# Convert the 'date' column to Date format
df$date <- ymd(df$date)

# Check if conversion worked
str(df$date)

# Filter for Advertising spending and analyze spending over time
advertising_spending <- df %>%
  filter(purpose_clean == "Advertising") %>%
  group_by(month = floor_date(date, "month")) %>%
  summarize(amount = sum(amount, na.rm = TRUE))
  

# Print results
print(advertising_spending)

```




**Q4.** Which committee has the larger payroll or salary spending so far? How many unique recipients of payroll/staff spending are there for each candidate?

**A4. Alsobrooks has the larger payroll/spending so far. Alsobrooks has three unique recipients while Hogan has zero. **

```{r}
# Ensure all candidates are included, even if they have no payroll spending
payroll_summary <- clean_md_senate_expenses %>%
  group_by(candidate) %>%  # Group before filtering
  summarise(
    total_payroll = sum(amount[purpose_clean %in% c("Payroll and Salary")], na.rm = TRUE),
    unique_recipients = n_distinct(organization_name[purpose_clean %in% c("Payroll and Salary")])
  )

# Print payroll spending by committee
print(payroll_summary)

```

**Q5** What's the most interesting story idea you've found in this data?

**A5 Why is there such a disparity between Alsobrooks and Hogan. I would want to look at why Alsobrooks has three unique recipients where she has spent over $1.26 million whereas Hogan has zero unique recipients  and have spent money. Additionally, I could look into campaign strategies.**
