---
title: "lab_09.Rmd"
author: "derek willis"
date: "2024-10-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   A Datawrapper account
-   Election data in CSV files in the `data` folder
-   Tidycensus

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
# load libraries here
library(tidyverse)
library(janitor)
library(tidycensus)
```

## Questions

**Q1.** You're interested in seeing how the methods of voting have changed in Maryland since 2018. Using the `md_votes_by_type.csv` that has data from the top of the ticket (governor in 2018 and 2022, president in 2020), make a chart showing the percentage of votes cast in each category - early voting, election day and mail - over the three years. You'll need to make those percentages in R and then export the CSV using `write_csv`, then make the chart in Datawrapper. Have a title, description and source that tells readers about the data. The title and description should highlight the most newsworthy or important aspect.

HINT: In Datawrapper's "Check and Describe" tab, you can click on a column to get the option of "hiding" it in any visualization.

What do you think the story here is? Put the answer below along with the URL of your chart.

**A1: The story here is that the 2020 election shifted the norms of voting methods, especially when it came to voting by mail. But this trend did not remain consistent during the 2022 election. Mail in voting dropped down to 19%. Election day voting surged to nearly 60% though as more people returned to the polls to vote in person. Early voting however has remained fairly consistent.  https://datawrapper.dwcdn.net/6ERC9/1/**


```{r}
md_votes <- read_csv("data/md_votes_by_type.csv") |>
  clean_names()

md_votes_pct <- md_votes |>
  mutate(
    early = early / total * 100,
    election_day = election_day / total * 100,
    mail = mail / total * 100,)

# Export for Datawrapper
write_csv(md_votes_pct, "data/md_votes_by_type_pct.csv")

```

**Q2.** You've been assigned to track early-voting turnout in Maryland at the county level. To do that, use [data from the state](https://elections.maryland.gov/press_room/2024_stats/PG24/EarlyVoting%20RAW%20data.csv). YOU DO NOT NEED TO LOAD THIS FILE. I've created several CSV files from it; each has a row for each day, with the counties as columns, so you can track early voting turnout in certain categories. The available files in the data folder are:

* ev_totals_18_24.csv - early voting turnout by the 18-24 age group
* ev_totals_65_up.csv - early voting turnout by the 65+ age group
* ev_totals_rep.csv - early voting turnout by Republicans
* ev_totals_dem.csv - early voting turnout by Democrats
* ev_totals_una.csv - early voting turnout by unaffiliated voters
* ev_totals_women.csv - early voting turnout by women
* ev_totals_men.csv - early voting turnout by men

Which file you use to make your chart, and what kind of chart is up to you (but it MUST be a chart, not a map or table), so experiment to see what works best. Since we have data over time, you might start with a line chart, but you might not end up there. You can display all counties, or just some of them - to limit the number of counties, you'll need to write code to select only the columns that you need and then use `write_csv()` to export the dataframe to a CSV file that you'll use in Datawrapper.

Your chart MUST: 

\* Have a title, description and source that tells readers about the data. The title and description should highlight the most newsworthy or important aspect. 
\* Easily readable labels and values 
\* Be published
\* Put the question you chose to answer below along, with the URL of your chart.

Tell me below why you picked the CSV you did. If you made any changes to the data, put that code below.

**A2:I picked the early voting totals csv file for women because I thought it would be interesting to look at how many women in various counties turn out to vote early and because women are becoming a key voting role and their votes can say a lot about their engagement in certain elections. I also picked this category because I am from Anne Arundel County and that was in the top five counties for most early votes by women. Below you will see that I modified my code because I tried to put all the data on a chart but it just became an eyesore to look at so I selected the top five counties that had the most early votes by women. Here is the link to the line chart: https://datawrapper.dwcdn.net/eohvM/2/** 

```{r}
# Load and clean the dataset
ev_women <- read_csv("data/ev_totals_women.csv") |>
  clean_names()

# Select top 5 counties + day
ev_women_top5 <- ev_women |>
  select(day, prince_georges, montgomery, baltimore_county,
         anne_arundel, howard)

# Export for Datawrapper
write_csv(ev_women_top5, "data/ev_women_top5_counties.csv")

```

**Q3.** After reading the recent CNN story about fundraising called "[How elderly dementia patients are unwittingly fueling political campaigns](https://www.cnn.com/interactive/2024/10/politics/political-fundraising-elderly-election-invs-dg/)," you decide to dig into the habits of Maryland political donors who give money via ActBlue and WinRed.

You want to show Maryland's political geography - where those donors live - and the frequency of donations. To do that, you'll load in two CSV files: `md_winred_zip_totals.csv` and `md_actblue_zip_totals.csv`.

You also will need to get population by zip code for Maryland - we did a similar thing in a previous lab, but this time you want the total population as your variable. Use tidycensus to do that, and save that dataframe, then join it to each of your contribution dataframes - you will need to make sure that the zip codes you're joining on are the same datatype.

In each dataframe, use the _number of donations_ to calculate a per capita rate of the number of donations per 1,000 population for each zip code using `mutate()`. Calculating a rate works like this: it's the thing you're counting divided by the population times the "per" number.

Then export each dataframe using `write_csv` and make two (2) choropleth maps in Datawrapper. For each map, you will need to:

-   Search for Maryland zip codes (use the 2020 data) as the base map
-   Upload the CSV file
-   Change the tooltip to show both the rate of donations and the total amount (you can put them side by side or put one under the other, as we did in the pre_lab). Make sure to properly format both numbers to make them readable.
-   Have a title, description and source that tells readers about the data (the source is the Federal Election Commission, and the data covers Jan. 1, 2023 through Oct. 16, 2024). The title and description should highlight the most newsworthy or important aspect.

Compare the two maps. What do you think are the most interesting story ideas in both, and what would you want to know more about? Put the answer below along with the URLs of your maps.

**A3: While both WinRed and ActBlue have the heaviest donations centered around the D.C. area and its surroundings, I was almost surprised by this. Many people probably expect the D.C. area to not have many donations by WinRed and would most likely assume that most of their donations would come from Western Maryland. Zip code 20854 belongs to Potomac, MD and they had a total of $944,799 donated by WinRed however, that doesn't begin to compare to the $4,750,407 that was donated by ActBlue in that same area. Keep in mind this wasn't ActBlue's largest total donation, that came from Chevy Chase, MD, with $9,090,681 donated. It's interesting to see how both WinRed and ActBlue both rely on the same areas for its donations. One thing I would like to know more about is who are some of the top donaors for WinRed and ActBlue and possibly look at the key areas that I pointed out for each one. Here are my links for each map: WinRed datawrapper link: https://datawrapper.dwcdn.net/hCnPa/4/ ActBlue datawrapper link: https://datawrapper.dwcdn.net/Npxog/1/** 

```{r}
winred_zip <- read_csv("data/md_winred_zip_totals.csv")
actblue_zip <- read_csv("data/md_actblue_zip_totals.csv")

# Get total population by ZIP
pop_data <- get_acs(
  geography = "zip code tabulation area",
  variables = "B01003_001",
  year = 2020,
  survey = "acs5"
) |> 
  mutate(GEOID = as.character(GEOID)) |> 
  rename(population = estimate)

# Prepare and join WinRed
winred_pop <- winred_zip |> 
  rename(GEOID = zip) |> 
  mutate(GEOID = as.character(GEOID)) |> 
  left_join(pop_data, by = "GEOID") |> 
  mutate(donations_percapita = (donations / population) * 1000)

# Prepare and join ActBlue
actblue_pop <- actblue_zip |> 
  rename(GEOID = zip) |> 
  mutate(GEOID = as.character(GEOID)) |> 
  left_join(pop_data, by = "GEOID") |> 
  mutate(donations_percapita = (donations / population) * 1000)

# Export to CSV
write_csv(winred_pop, "data/winred.csv")
write_csv(actblue_pop, "data/actblue.csv")

```
